---
title: "Ranked Choice Voting"
author: "Patrick Kelly"
date: "12/6/2018"
output: pdf_document
---
#Ranked Choice Voting
##Intro to RCV in Maine

###Ranked Choice Voting Basics

http://www.lwvme.org/files/RCV_Basics.pdf

### Ranked-Choice Voting in Maine (Video)

https://youtu.be/vp1APVk7SQk)

###Here are the Rules

1. Voters rank their choices in order of preference from left to right.
2. Two choices in the same column is an overvote, and the vote is removed.
3. An empty column is an undervote.
4. If the column to the right of an undervote has a candidate, that vote is moved to the column on its left.
5. If there are undervotes in two consecutive columns, then then that vote is removed.
6. If there are overrvotes in the first choices, remove them and tally the totals, including any undervotes.
7. If there are undervotes and any of them have a valid vote in the second column, then move that vote to the first column.
8. If the first and second columns are both undervotes, then that vote is removed.
9. Now tally the first choice votes.
10. If the leader does not have at least 50% +1 of the votes, proceed to Round 1.
11. And if necessary, continue to additional rounds until one of the two leading candidates has a majority of the valid votes.

# Load the data
### These data include overvotes and undervotes, as happens with real world data.

### Load data from Github

```{r}
path <- "https://raw.github.com/damonzon/Rank_Choice_Voting/master/RC_Example3.csv"
example <- read.csv(path)
```

\pagebreak

```{r}
suppressMessages(library(dplyr))
example
```

\pagebreak

```{r}
table(example$First)
round(prop.table(table(example$First)),3)

```
### There are 2 overvotes to be removed.

```{r}
example <- filter(example,First != "Overvote")
table(example$First)
round(prop.table(table(example$First)),3)
```
### Check the undervote to see if a name is in column 2.

```{r}
Continuing_Votes <- filter (example,First != "Undervote")
nrow(Continuing_Votes)
undervote <- filter(example,First == "Undervote")
undervote
```
###We see 2 "Blue" in column two, so they are moved left to column one.

\pagebreak

```{r}
uv2 <- function(x){ 
undervote2 <<- undervote
undervote2$First <<- undervote2$Second
undervote2$Second <<- undervote2$Third
undervote2$Third <<- undervote2$Fourth
undervote2$fourth <<- undervote2$"Undervote"
table(undervote2$First)
}
uv2(undervote)
```

```{r}
undervote3 <- filter(undervote2,First != "Undervote")
undervote3 <- filter (undervote3,First != "Overvote")
table(undervote3$First)
```

##Add these votes to the first choice Continuing Votes 

```{r}
data <- rbind(Continuing_Votes,undervote3)
rm(Continuing_Votes,undervote,undervote2,undervote3)
table(data$First)
round(prop.table(table(data$First)),3)
nrow(data)
```
##Yellow leads in first choice votes with 33.3%. Since this is not over 50%, we proceed by droping Orange and proceeding to the Round 1.
##Remove Orange, the category with the fewest votes (5).

\pagebreak

```{r}
data2 <- filter(data,First != "Orange")
table(data2$First)
orange <- filter(data,First == "Orange")
orange2 <- orange
orange2$First <- orange2$Second
orange2$Second <- orange2$Third
orange$Third <- orange2$Fourth
orange2$Fourth <- "Undervote"
table(orange2$First)
data3 <- filter(orange2,First != "Undervote")
table(data3$First)
undervote <- filter(orange2,First == "Undervote")
uv2(undervote)
```

```{r}
undervote3 <- filter(undervote2,First != "Undervote")
undervote3 <- filter(undervote3,First != "Orange")
undervote3<- filter (undervote3,First != "Overvote")
table(undervote3$First)
data4 <- rbind(data3,undervote3)
data <- rbind(data2,data4)
rm(data2, data3,data4)
rm(orange, orange2)
rm(undervote,undervote2,undervote3)
nrow(data)
table(data$First)
round(prop.table(table(data$First)),3)
```

##Blue is now leading with 11 votes (36.7%), but not a majority. So we will now remove Purple and proceed to Round 2.

```{r}
data2 <- filter(data,First != "Purple")
table(data2$First)
purple <- filter(data,First == "Purple")
purple2 <- purple
purple2$First <- purple2$Second
purple2$Second <- purple2$Third
purple2$Third <- purple2$Fourth
purple2$Fourth <- "Undervote"
table(purple2$First)
```

```{r}
purple2<- filter (purple2,First != "Purple")
purple2<- filter (purple2,First != "Overvote")
table(purple2$First)
data3 <- filter(purple2,First != "Undervote")
table(data3$First)
undervote <- filter(purple2,First == "Undervote")
uv2(undervote)
```

# Remove Orange and any overvote and undervote

```{r}
undervote2<- filter (undervote2,First != "Orange")
undervote2<- filter (undervote2,First != "Overvote")
undervote2 <- filter(undervote2,First != "Undervote")
table(undervote2$First)
```

```{r}
table(data3$First)
data4 <- rbind(data3,undervote2)
table(data4$First)
data <- rbind(data2,data4)
table(data$First)
round(prop.table(table(data$First)),3)
```

### Now remove Orange again and proceed to Round 3.

```{r}
data2 <- filter(data,First != "Orange")
table(data2$First)
orange <- filter(data,First == "Orange")
nrow(orange)
orange2 <- orange
orange2$First <- orange2$Second
orange2$Second <- orange2$Third
orange2$Third <- orange2$Fourth
orange2$Fourth <- "Undervote"
table(orange2$First)
orange2<- filter (orange2,First != "Orange")
orange2<- filter (orange2,First != "Overvote")
table(orange2$First)
data3 <- filter(orange2,First != "Undervote")
table(data3$First)
undervote <- filter(orange2,First == "Undervote")
uv2(undervote)
```

```{r}
undervote3 <- filter(undervote2,First != "Undervote")
undervote3 <- filter(undervote3,First != "Orange")
undervote3<- filter (undervote3,First != "Overvote")
table(undervote3$First)
data4 <- rbind(data3,undervote3)
table(data4$First)
data <- rbind(data2,data4)
table(data$First)
round(prop.table(table(data$First)),3)
rm(data2, data3,data4)
rm(orange, orange2)
rm(undervote,undervote2,undervote3)
#table(data$first)
nrow(data)
```

###Finally Blue wins with a majority: 14 votes (53.8%) compared to 12 votes (46.2%) for Yellow.

http://www.mainepublic.org/post/its-just-math-mainers-recreate-ranked-choice-voting-results-test-process

